\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-08-05 manuscript/Rmd/draft.Rmd
Plenty to edit still

abstract: |
  The purpose of this paper is to investigate the controversial competition format and scoring system of sport climbing, which is one of the sports making its debut on the Olympics stage at Tokyo 2020. Climbing Olympians will participate in three disciplines: speed climbing, bouldering, and lead climbing for a single set of medals, and a multiplied score of climbers' rankings in all three disciplines will be used to determine the overall standings. This work finds great evidence from historical data and simulations that speed climbing speed climbing should be separated from the combined format and have its own medals. Furthermore, it is shown that the product of ranks scoring system violates the independence of irrelevant alternatives property.

@@

The 8 climbers with the lowest score (i.e. lowest product of ranks across the three disciplines) in the qualification round advance to the finals where they once again compete in all three disciplines. The final score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round, with the main difference being that there are only 8 competitors in the final as opposed to 20 in the qualification round.  The climbers with the lowest, second lowest, and third lowest product of ranks in the final wins the gold, silver, and bronze medal, respectively. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

@@

To the best of our knowledge, we know of no sporting event, team or individual, that uses the product of ranks to determine an overall rankings.  There are examples of team sports that use the sum of ranks to determine the winning team such as cross country, where the team with the lowest sum of ranks of the top five runners is determined to be the winner. For more detail on rank sum scoring, see @hammond2007 and @boudreau2018. In addition, some individual sports such as the decathlon and heptathlon rely on a sum of scores from the ten or seven events, however, these scores are not determined based on the ranks of the competitors.  That is, a decathlete's score is entirely based upon their times, distances, and heights, and their overall score will be exactly the same if the times, distances, and heights remain the same regardless of the performance of other individuals.  Full details of decathlon scoring can be found in @westera2006.

<!-- 
There are other individual competitions consisted of several events combined that do base their scoring on ranks, such as crossfit competitions.  In each event, points are earned based on the competitor's rank in the event based on a scoring table. (https://games.crossfit.com/announcement/12940), and a competitors final score is based on the sum of their scores across all the events.

Paragraph on scoring methods in other sports: Soccer points: @haugen2008; Cross country scoring: @hammond2007, @boudreau2018; Decathlon scoring: @westera2006

As a side note, rank product, genes 
-->

The manuscript is outlined as follows. We first begin with some descriptions of the data and methods in Section \ref{data-and-methods}. Our analyses and results are then presented in Sections \ref{simulations} and \ref{data-analysis}. Finally, in Section \ref{conclusion-and-discussion}, we summarize up our main findings and provide a discussion to close out the paper.

@@

Additionally, we rely historical results from major climbing competitions in recent years. We collected data on climbing contests that took place between 2018 and 2020, where the combined format was used to determine the scores and ranks of climbers. The events include the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Youth Olympics. Data were obtained from various sources, including the event websites, Wikipedia, and the International Federation of Sport Climbing (IFSC). The main attributes of our datasets are the name and nationality of the climbers; bib number (for some competitions); the finishing place of climbers in speed climbing, bouldering, and lead climbing; the total score (which equals the product of event ranks); and the final rank. We utilize this data to compute the correlations between the event ranks and final table position, as well as to look at how often the final orderings change if one athlete is dropped and the ranks for each discipline are re-computed.

Paragraph on IIA/Social Choice

Suppose we have 3 people A, B, and C participating in a competition. If A finishes in the first place and C is later disqualified and removed, A should still win. If the original winner (A) loses the modified competition (with C removed), then the Independence of Irrelevant Alternatives has been violated.

Economics, psychology, probability theory. Connect with cross country in previous section

First mentioned by Arrow (1951) https://cowles.yale.edu/sites/default/files/files/pub/mon/m12-all.pdf

@@

Ray (1973) https://www.jstor.org/stable/1913820

@@

In addition, the average score for qualification positions 1 to 8 are displayed in Table X. We notice that on average, the minimum score that one should aim for in order to move on to the final round is about 434 (for 8th rank).

@@

```{r, fig.cap="Probability of finishing at every rank given", fig.width = 6.4, fig.height = 3.2}
win_any_qual %>% 
  mutate(type = "Win Any") %>% 
  bind_rows(mutate(win_first_qual, type = "Win First")) %>% 
  mutate(finalist = ifelse(rank %in% 1:8, "Yes", "No")) %>% 
  ggplot(aes(rank, weight = prob, fill = finalist)) +
  geom_bar() +
  geom_vline(xintercept = 8.5) +
  scale_x_continuous() +
  labs(x = "Rank",
       y = "Probability",
       fill = "Finalist") +
  facet_wrap(~ type)
```

@@

Regarding the finals, a climber is very likely to finish in the top 3 and hence earn a medal if they win the first event (83.03% chance) or any event (85.01% chance). Furthermore, according to our final simulations, in order to obtain a climbing medal, the average scores (rounded down) that put an athlete in position to receive gold, silver, and bronze medals are 9, 20, and 33, respectively (see Table 3).

@@

```{r, fig.cap="Probability of finishing at every rank given", fig.width = 6.4, fig.height = 3.2}
win_any_final %>% 
  mutate(type = "Win Any") %>% 
  bind_rows(mutate(win_first_final, type = "Win First")) %>% 
  mutate(medalist = ifelse(rank %in% 1:3, "Yes", "No")) %>% 
  ggplot(aes(rank, weight = prob, fill = medalist)) +
  geom_bar() +
  geom_vline(xintercept = 3.5) +
  scale_x_continuous() +
  labs(x = "Rank",
       y = "Probability",
       fill = "Medalist") +
  facet_wrap(~ type)
```

@@

sum_dat1 <- reorder_all_above %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n()
            , "Avg. Probability" = mean(displ_prob)
            , "Median Probability" = median(displ_prob)
            , "St. Deviation" = sd(displ_prob)
            )

kable(sum_dat1)
```

@@

reorder_dat2 <- lapply(9:20
                  , function(x) {
                    data.frame(dropped_rank = x
                               , displ_prob = pct_displace_finalists[[x-8]]
                               )
                    }
                  ) %>%
  do.call(rbind, .)

sum_dat2 <- reorder_dat2 %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n()
            , "Avg. Probability" = mean(displ_prob)
            , "Median Probability" = median(displ_prob)
            , "St. Deviation" = sd(displ_prob)
            )

kable(sum_dat2)

@@

displace_probs %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n()
            , "Avg. Probability" = mean(displ_prob)
            , "Median Probability" = median(displ_prob)
            , "St. Deviation" = sd(displ_prob)
            ) %>%
 kable()

@@

Figure 3 shows the modified versions of the rankings after each ranked climber is excluded for both gender events. We have clear evidence from this plot that removing a single climber changes the rankings drastically, especially order of medalists. One particular interesting case is where an athlete's position change when someone who originally finished behind them drops out. This situation is illustrated by panel 5 of the women's competition, where the fifth-place climber, Krasovskaia, was excluded; and Meul, whose actual final rank was fourth, moved up to the second spot and claim the silver medal.

@@

```{r}
mf_gg <- drop_rerank(mf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 4) | 
             rank_drop == 1 & rank %in% 4:5 |
              rank_drop == 2 & rank %in% c(1, 5) |
                rank_drop %in% c(3, 5) & rank %in% c(1, 2, 5) |
                  rank_drop == 6 & rank %in% 1:3, 
           "no", "yes")) %>% 
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(1, 2, 3, 5, 6)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2) +
  expand_limits(y = 102, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(x = "Men", y = "") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())

>>>>

abstract: |
  The purpose of this paper is to investigate the controversial competition format and scoring system of sport climbing, which is one of the sports making its debut on the Olympics stage at Tokyo 2020. Climbing Olympians are required to participate in three disciplines: speed climbing, bouldering, and lead climbing for only a single set of medals, and a multiplied score of climbers' rankings in all three disciplines will be used to determine the overall standings. This work finds great evidence from historical data and simulations that the current competition format is putting speed climbers at a disadvantage. Furthermore, it is shown that the product of ranks scoring system violates the independence of irrelevant alternatives property.

@@

The 8 climbers with the lowest score (in terms of product of ranks across the three disciplines) in the qualification round advance to the finals where they once again compete in all three disciplines. The final score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round, with the main difference being that there are only 8 competitors in the final as opposed to 20 in the qualification round.  The climbers with the lowest, second lowest, and third lowest product of ranks in the final wins the gold, silver, and bronze medal, respectively. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

@@

To the best of our knowledge, we know of no sporting event, team or individual, that uses the product of ranks to determine an overall rankings.  There are examples of team sports that use the sum of ranks to determine the winning team such as cross country, where the team with the lowest sum of ranks of the top five runners is determined to be the winner [@hammond2007; @boudreau2018]. In addition, some individual sports such as the decathlon and heptathlon rely on a sum of scores from the ten or seven events, however, these scores are not determined based on the ranks of the competitors.  That is, a decathlete's score is entirely based upon their times, distances, and heights, and their overall score will be exactly the same if the times, distances, and heights remain the same regardless of the performance of other individuals [@westera2006]. Furthermore, there are other individual competitions consisted of several events combined that do base their scoring on ranks, such as crossfit competitions.  In each event, points are earned based on the competitor's rank in the event based on a scoring table, and a competitors final score is based on the sum of their scores across all the events [@crossfit2021].

<!--
Paragraph on scoring methods in other sports: Soccer points: @haugen2008; Cross country scoring: @hammond2007, @boudreau2018; Decathlon scoring: @westera2006
As a side note, rank product, genes

In particular, we show how this format is putting speed climbers at a disadvantage, as well as address the 
-->

In this paper, we attempt to use simple statistical analysis to address the limitations of sport climbing's combined competition format and ranking system.  The manuscript is outlined as follows. We first begin with some descriptions of the data and methods in Section \ref{data-and-methods}. Our analyses and results are then presented in Sections \ref{simulations} and \ref{data-analysis}. Finally, in Section \ref{conclusion-and-discussion}, we summarize up our main findings and provide a discussion to close out the paper.

@@

Additionally, we rely historical results from major climbing competitions in recent years. We collected data on climbing contests that took place between 2018 and 2020, where the combined format was used to determine the scores and ranks of climbers. The events include the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Youth Olympics. Data were obtained from various sources, including the event websites, Wikipedia, and the International Federation of Sport Climbing (IFSC). The main attributes of our datasets are the name and nationality of the climbers; bib number (for some competitions); the finishing place of climbers in speed climbing, bouldering, and lead climbing; the total score (which equals the product of event ranks); and the final rank. We utilize this data to compute the correlations between the event ranks and final table position, as well as to look at how often the final orderings change if one athlete is dropped and the ranks for each discipline as well as the total score are re-calculated.

Since we are interested in applying the "leave-one-climber-out" type of analysis to sport climbing, there is a connection between this situation and the idea of independence of irrelevant alternatives (IIA). The IIA criterion is a property of a voting system which states that after a winner is determined, if one of the losing candidates drops out and the votes are recounted, there should not be a change in the winner. First mentioned by @arrow1951, the IIA condition is also known as Luce's choice axiom [@luce1959] in probability theory, and it has had a number of applications in the fields of decision theory, economics, and psychology over the years.

We notice a link between the concept of IIA and the topic of ranking system in sports. As an illustration, suppose we have 3 players A, B, and C participating in a competition. If A finishes in the first place and C is later disqualified and removed, A should still win. If the original winner (A) loses the modified competition (with C removed), then the Independence of Irrelevant Alternatives has been violated. For our particular case, this sort of analysis can be helpful in examining the results of a climbing competition, specifically how a disqualification can affect the standings of medalists in the final round.

<!--
First mentioned by Arrow (1951) https://cowles.yale.edu/sites/default/files/files/pub/mon/m12-all.pdf

@@

Ray (1973) https://www.jstor.org/stable/1913820
-->

@@

In addition, the average score for qualification positions 1 to 8 are displayed in Table 3. We observe that on average, the minimum score that one should aim for in order to move on to the final round is about 434 (for 8th rank).

@@

```{r, fig.cap="Distribution plots of qualification ranks, given 1) winning any event and 2) winning first event", fig.width = 6.4, fig.height = 3.2}
win_any_qual %>% 
  mutate(type = "Win Any") %>% 
  bind_rows(mutate(win_first_qual, type = "Win First")) %>% 
  mutate(finalist = ifelse(rank %in% 1:8, "Yes", "No")) %>% 
  ggplot(aes(rank, weight = prob, fill = finalist)) +
  geom_bar() +
  geom_vline(xintercept = 8.5) +
  scale_x_continuous(breaks = 1:11) +
  labs(x = "Rank",
       y = "Probability",
       fill = "Finalist") +
  facet_wrap(~ type)
```

@@

Regarding the finals, Table 4 indicates that a climber is very likely to finish in the top 3 and hence earn a medal if they win the first event (83.03% chance) or any event (85.01% chance). Furthermore, according to our final simulations, in order to obtain a climbing medal, the average scores (rounded down) that put an athlete in position to receive gold, silver, and bronze medals are 9, 20, and 33, respectively (see Table 5).

@@

```{r, fig.cap="Distribution plots of final ranks, given 1) winning any event and 2) wiining first event", fig.width = 6.4, fig.height = 3.2}
win_any_final %>% 
  mutate(type = "Win Any") %>% 
  bind_rows(mutate(win_first_final, type = "Win First")) %>% 
  mutate(medalist = ifelse(rank %in% 1:3, "Yes", "No")) %>% 
  ggplot(aes(rank, weight = prob, fill = medalist)) +
  geom_bar() +
  geom_vline(xintercept = 3.5) +
  scale_x_continuous(breaks = 1:7) +
  labs(x = "Rank",
       y = "Probability",
       fill = "Medalist") +
  facet_wrap(~ type)
```

@@

reorder_all_above %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n(), 
            "Avg. Probability" = mean(displ_prob), 
            "Median Probability" = median(displ_prob), 
            "St. Deviation" = sd(displ_prob)) %>% 
  kable(booktabs = TRUE)
```

@@

reorder_dat2 <- lapply(9:20, 
                       function(x) {
                         data.frame(dropped_rank = x, 
                                    displ_prob = pct_displace_finalists[[x - 8]])}) %>%
  bind_rows()

reorder_dat2 %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n(), 
            "Avg. Probability" = mean(displ_prob), 
            "Median Probability" = median(displ_prob),
            "St. Deviation" = sd(displ_prob)) %>% 
  kable(booktabs = TRUE)
```

@@

displace_probs %>%
  group_by(dropped_rank) %>%
  rename("Dropped Rank" = dropped_rank) %>%
  summarize("Simulations" = n(), 
            "Avg. Probability" = mean(displ_prob), 
            "Median Probability" = median(displ_prob), 
            "St. Deviation" = sd(displ_prob)) %>%
 kable(booktabs = TRUE)

@@

Figure 3 shows the modified versions of the rankings after each ranked climber is excluded for both gender events. We have clear evidence from this plot that removing a single climber changes the rankings drastically, especially order of medalists. One particular interesting case is where an athlete's position change when someone who originally finished behind them drops out. This situation is illustrated by panel 5 of the women's competition, where the fifth-place climber, Krasovskaia, was excluded; and Meul, whose actual final rank was fourth, moved up to the second spot and claim the silver medal. Furthermore, this implies that the ranking system of Olympics sport climbing violates the IIA criterion.

@@

```{r}
mf_gg <- drop_rerank(mf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 4) | 
             rank_drop == 1 & rank %in% 4:5 |
              rank_drop == 2 & rank %in% c(1, 5) |
                rank_drop %in% c(3, 5) & rank %in% c(1, 2, 5) |
                  rank_drop == 6 & rank %in% 1:3, 
           "no", "yes")) %>% 
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(1, 2, 3, 5, 6)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2) +
  expand_limits(y = 104, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(x = "Men", y = "") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())

### 2021-08-05 manuscript/Rmd/draft.tex
Plenty to edit still

\bigskip
\begin{abstract}
The purpose of this paper is to investigate the controversial
competition format and scoring system of sport climbing, which is one of
the sports making its debut on the Olympics stage at Tokyo 2020.
Climbing Olympians will participate in three disciplines: speed
climbing, bouldering, and lead climbing for a single set of medals, and
a multiplied score of climbers' rankings in all three disciplines will
be used to determine the overall standings. This work finds great
evidence from historical data and simulations that speed climbing speed
climbing should be separated from the combined format and have its own
medals. Furthermore, it is shown that the product of ranks scoring
system violates the independence of irrelevant alternatives property.
\end{abstract}

@@

The 8 climbers with the lowest score (i.e.~lowest product of ranks
across the three disciplines) in the qualification round advance to the
finals where they once again compete in all three disciplines. The final
score for each person in the final stage is determined by multiplying
their ranks in each discipline, similar to the qualification round, with
the main difference being that there are only 8 competitors in the final
as opposed to 20 in the qualification round. The climbers with the
lowest, second lowest, and third lowest product of ranks in the final
wins the gold, silver, and bronze medal, respectively. This type of
scoring system heavily rewards high finishes and relatively ignores poor
finishes. For instance, if climber A finished 1st, 20th, and 20th and
climber B finished 10th, 10th, and 10th, climber B would have a score of
1000 whereas climber A would have a much better score of 400, despite
finishing last in 2 out of 3 of the events.

To the best of our knowledge, we know of no sporting event, team or
individual, that uses the product of ranks to determine an overall
rankings. There are examples of team sports that use the sum of ranks to
determine the winning team such as cross country, where the team with
the lowest sum of ranks of the top five runners is determined to be the
winner. For more detail on rank sum scoring, see \citet{hammond2007} and
\citet{boudreau2018}. In addition, some individual sports such as the
decathlon and heptathlon rely on a sum of scores from the ten or seven
events, however, these scores are not determined based on the ranks of
the competitors. That is, a decathlete's score is entirely based upon
their times, distances, and heights, and their overall score will be
exactly the same if the times, distances, and heights remain the same
regardless of the performance of other individuals. Full details of
decathlon scoring can be found in \citet{westera2006}.

The manuscript is outlined as follows. We first begin with some
descriptions of the data and methods in Section \ref{data-and-methods}.
Our analyses and results are then presented in Sections
\ref{simulations} and \ref{data-analysis}. Finally, in Section
\ref{conclusion-and-discussion}, we summarize up our main findings and
provide a discussion to close out the paper.

@@

Additionally, we rely historical results from major climbing
competitions in recent years. We collected data on climbing contests
that took place between 2018 and 2020, where the combined format was
used to determine the scores and ranks of climbers. The events include
the 2020 Continental Championships of Europe, Africa, Oceania,
Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and
2018 Youth Olympics. Data were obtained from various sources, including
the event websites, Wikipedia, and the International Federation of Sport
Climbing (IFSC). The main attributes of our datasets are the name and
nationality of the climbers; bib number (for some competitions); the
finishing place of climbers in speed climbing, bouldering, and lead
climbing; the total score (which equals the product of event ranks); and
the final rank. We utilize this data to compute the correlations between
the event ranks and final table position, as well as to look at how
often the final orderings change if one athlete is dropped and the ranks
for each discipline are re-computed.

Paragraph on IIA/Social Choice

Suppose we have 3 people A, B, and C participating in a competition. If
A finishes in the first place and C is later disqualified and removed, A
should still win. If the original winner (A) loses the modified
competition (with C removed), then the Independence of Irrelevant
Alternatives has been violated.

Economics, psychology, probability theory. Connect with cross country in
previous section

First mentioned by Arrow (1951)
\url{https://cowles.yale.edu/sites/default/files/files/pub/mon/m12-all.pdf}

Luce (1959)
\url{http://www.scholarpedia.org/article/Luce\%27s_choice_axiom}

and Luce (1977)
\url{https://www.imbs.uci.edu/files/personnel/luce/pre1990/1977/Luce_JMP_1977a.pdf}

Ray (1973) \url{https://www.jstor.org/stable/1913820}

@@

\caption{Probability of finishing at every rank given}\label{fig:unnamed-chunk-7}
\end{figure}

@@

Regarding the finals, a climber is very likely to finish in the top 3
and hence earn a medal if they win the first event (83.03\% chance) or
any event (85.01\% chance). Furthermore, according to our final
simulations, in order to obtain a climbing medal, the average scores
(rounded down) that put an athlete in position to receive gold, silver,
and bronze medals are 9, 20, and 33, respectively (see Table 3).

@@

\caption{Probability of finishing at every rank given}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\begin{tabular}{r|r|r|r|r}
\hline
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\hline
4 & 100 & 0.25869 & 0.2590 & 0.0145433\\
\hline
5 & 100 & 0.34032 & 0.3405 & 0.0151263\\
\hline
6 & 100 & 0.38012 & 0.3820 & 0.0152430\\
\hline
7 & 100 & 0.40636 & 0.4060 & 0.0149582\\
\hline
8 & 100 & 0.39439 & 0.3930 & 0.0151023\\
\hline
\end{tabular}

@@

\begin{tabular}{r|r|r|r|r}
\hline
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\hline
9 & 100 & 0.0505 & 0.050 & 0.0217597\\
\hline
10 & 100 & 0.1538 & 0.150 & 0.0318386\\
\hline
11 & 100 & 0.1603 & 0.160 & 0.0343616\\
\hline
12 & 100 & 0.1416 & 0.140 & 0.0353259\\
\hline
13 & 100 & 0.1365 & 0.140 & 0.0336162\\
\hline
14 & 100 & 0.1217 & 0.120 & 0.0344965\\
\hline
15 & 100 & 0.1213 & 0.120 & 0.0308353\\
\hline
16 & 100 & 0.1159 & 0.120 & 0.0306197\\
\hline
17 & 100 & 0.1086 & 0.105 & 0.0314312\\
\hline
18 & 100 & 0.1026 & 0.100 & 0.0329867\\
\hline
19 & 100 & 0.0886 & 0.090 & 0.0291641\\
\hline
20 & 100 & 0.0700 & 0.070 & 0.0258590\\
\hline
\end{tabular}

@@

\begin{tabular}{r|r|r|r|r}
\hline
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\hline
4 & 100 & 0.25872 & 0.2580 & 0.0137106\\
\hline
5 & 100 & 0.31988 & 0.3190 & 0.0161416\\
\hline
6 & 100 & 0.28741 & 0.2870 & 0.0127969\\
\hline
7 & 100 & 0.25631 & 0.2545 & 0.0145301\\
\hline
8 & 100 & 0.20631 & 0.2050 & 0.0120283\\
\hline
\end{tabular}

>>>>

\bigskip
\begin{abstract}
The purpose of this paper is to investigate the controversial
competition format and scoring system of sport climbing, which is one of
the sports making its debut on the Olympics stage at Tokyo 2020.
Climbing Olympians are required to participate in three disciplines:
speed climbing, bouldering, and lead climbing for only a single set of
medals, and a multiplied score of climbers' rankings in all three
disciplines will be used to determine the overall standings. This work
finds great evidence from historical data and simulations that the
current competition format is putting speed climbers at a disadvantage.
Furthermore, it is shown that the product of ranks scoring system
violates the independence of irrelevant alternatives property.
\end{abstract}

@@

The 8 climbers with the lowest score (in terms of product of ranks
across the three disciplines) in the qualification round advance to the
finals where they once again compete in all three disciplines. The final
score for each person in the final stage is determined by multiplying
their ranks in each discipline, similar to the qualification round, with
the main difference being that there are only 8 competitors in the final
as opposed to 20 in the qualification round. The climbers with the
lowest, second lowest, and third lowest product of ranks in the final
wins the gold, silver, and bronze medal, respectively. This type of
scoring system heavily rewards high finishes and relatively ignores poor
finishes. For instance, if climber A finished 1st, 20th, and 20th and
climber B finished 10th, 10th, and 10th, climber B would have a score of
1000 whereas climber A would have a much better score of 400, despite
finishing last in 2 out of 3 of the events.

To the best of our knowledge, we know of no sporting event, team or
individual, that uses the product of ranks to determine an overall
rankings. There are examples of team sports that use the sum of ranks to
determine the winning team such as cross country, where the team with
the lowest sum of ranks of the top five runners is determined to be the
winner \citep{hammond2007, boudreau2018}. In addition, some individual
sports such as the decathlon and heptathlon rely on a sum of scores from
the ten or seven events, however, these scores are not determined based
on the ranks of the competitors. That is, a decathlete's score is
entirely based upon their times, distances, and heights, and their
overall score will be exactly the same if the times, distances, and
heights remain the same regardless of the performance of other
individuals \citep{westera2006}. Furthermore, there are other individual
competitions consisted of several events combined that do base their
scoring on ranks, such as crossfit competitions. In each event, points
are earned based on the competitor's rank in the event based on a
scoring table, and a competitors final score is based on the sum of
their scores across all the events \citep{crossfit2021}.

In this paper, we attempt to use simple statistical analysis to address
the limitations of sport climbing's combined competition format and
ranking system. The manuscript is outlined as follows. We first begin
with some descriptions of the data and methods in Section
\ref{data-and-methods}. Our analyses and results are then presented in
Sections \ref{simulations} and \ref{data-analysis}. Finally, in Section
\ref{conclusion-and-discussion}, we summarize up our main findings and
provide a discussion to close out the paper.

@@

Additionally, we rely historical results from major climbing
competitions in recent years. We collected data on climbing contests
that took place between 2018 and 2020, where the combined format was
used to determine the scores and ranks of climbers. The events include
the 2020 Continental Championships of Europe, Africa, Oceania,
Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and
2018 Youth Olympics. Data were obtained from various sources, including
the event websites, Wikipedia, and the International Federation of Sport
Climbing (IFSC). The main attributes of our datasets are the name and
nationality of the climbers; bib number (for some competitions); the
finishing place of climbers in speed climbing, bouldering, and lead
climbing; the total score (which equals the product of event ranks); and
the final rank. We utilize this data to compute the correlations between
the event ranks and final table position, as well as to look at how
often the final orderings change if one athlete is dropped and the ranks
for each discipline as well as the total score are re-calculated.

Since we are interested in applying the ``leave-one-climber-out'' type
of analysis to sport climbing, there is a connection between this
situation and the idea of independence of irrelevant alternatives (IIA).
The IIA criterion is a property of a voting system which states that
after a winner is determined, if one of the losing candidates drops out
and the votes are recounted, there should not be a change in the winner.
First mentioned by \citet{arrow1951}, the IIA condition is also known as
Luce's choice axiom \citep{luce1959} in probability theory, and it has
had a number of applications in the fields of decision theory,
economics, and psychology over the years.

We notice a link between the concept of IIA and the topic of ranking
system in sports. As an illustration, suppose we have 3 players A, B,
and C participating in a competition. If A finishes in the first place
and C is later disqualified and removed, A should still win. If the
original winner (A) loses the modified competition (with C removed),
then the Independence of Irrelevant Alternatives has been violated. For
our particular case, this sort of analysis can be helpful in examining
the results of a climbing competition, specifically how a
disqualification can affect the standings of medalists in the final
round.

@@

\caption{Distribution plots of qualification ranks, given 1) winning any event and 2) winning first event}\label{fig:unnamed-chunk-7}
\end{figure}

@@

Regarding the finals, Table 4 indicates that a climber is very likely to
finish in the top 3 and hence earn a medal if they win the first event
(83.03\% chance) or any event (85.01\% chance). Furthermore, according
to our final simulations, in order to obtain a climbing medal, the
average scores (rounded down) that put an athlete in position to receive
gold, silver, and bronze medals are 9, 20, and 33, respectively (see
Table 5).

@@

\caption{Distribution plots of final ranks, given 1) winning any event and 2) wiining first event}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\begin{tabular}{rrrrr}
\toprule
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\midrule
4 & 100 & 0.25869 & 0.2590 & 0.0145433\\
5 & 100 & 0.34032 & 0.3405 & 0.0151263\\
6 & 100 & 0.38012 & 0.3820 & 0.0152430\\
7 & 100 & 0.40636 & 0.4060 & 0.0149582\\
8 & 100 & 0.39439 & 0.3930 & 0.0151023\\
\bottomrule
\end{tabular}

@@

\begin{tabular}{rrrrr}
\toprule
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\midrule
9 & 100 & 0.0505 & 0.050 & 0.0217597\\
10 & 100 & 0.1538 & 0.150 & 0.0318386\\
11 & 100 & 0.1603 & 0.160 & 0.0343616\\
12 & 100 & 0.1416 & 0.140 & 0.0353259\\
13 & 100 & 0.1365 & 0.140 & 0.0336162\\
\addlinespace
14 & 100 & 0.1217 & 0.120 & 0.0344965\\
15 & 100 & 0.1213 & 0.120 & 0.0308353\\
16 & 100 & 0.1159 & 0.120 & 0.0306197\\
17 & 100 & 0.1086 & 0.105 & 0.0314312\\
18 & 100 & 0.1026 & 0.100 & 0.0329867\\
\addlinespace
19 & 100 & 0.0886 & 0.090 & 0.0291641\\
20 & 100 & 0.0700 & 0.070 & 0.0258590\\
\bottomrule
\end{tabular}

@@

\begin{tabular}{rrrrr}
\toprule
Dropped Rank & Simulations & Avg. Probability & Median Probability & St. Deviation\\
\midrule
4 & 100 & 0.25872 & 0.2580 & 0.0137106\\
5 & 100 & 0.31988 & 0.3190 & 0.0161416\\
6 & 100 & 0.28741 & 0.2870 & 0.0127969\\
7 & 100 & 0.25631 & 0.2545 & 0.0145301\\
8 & 100 & 0.20631 & 0.2050 & 0.0120283\\
\bottomrule
\end{tabular}
